Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-13 content/contribution/continuous-datalog.tex
Fixed minor formatting isues

Continuous Datalog is an extension of positive Datalog with
stream-processing semantics, with a view to accommodating the aims of
the latter within the well-defined model-theoretic semantics of the
former, which are shared with RIF-Core \cite{w3crifbld} and therefore
support reasoning at the expressivity of OWL 2 RL.  This comprises the
extension of Datalog programs to include sources of continuously
generated/reported facts, and the application of sliding window
semantics over those sources, thereby providing for the interpretation
of Continuous Datalog programs as sequences of instantaneous states
that are each valid positive Datalog programs.  As such, these states
each inherit the semantics of positive Datalog.  Furthermore, it
involves the continuous derivation of entailments from this sequence
of instantaneous states as a set of streams of facts; these streams
share a data model with the streams of continuous queries, as defined
by the stream processing community, thereby allowing the integration
of Continuous Datalog programs into stream processing pipelines.

@@

\begin{definition}[Streams]
\labeldef{continuous datalog: stream}
A {\em stream} is a sequence of fact instances ordered
by their instants.
%FIXME: does this mean that a stream never contains duplicates?
%; in conjunction with the distinction between
%multiple instances of a fact by their timestamp, this definition justifies
%Axiom~\ref{axiom:continuous datalog: window disjointness}.
\end{definition}

@@

From this description of instance entailment,
Axiom~\ref{axiom:continuous datalog: entailment disjointness} may be
derived:
\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
  If two sliding windows over the same stream $S$ cover non-overlapping time periods,
then a set $E^T_1$ of fact instances entailed by ${\langle R, W^{S,x}_i \rangle}$
is necessarily disjoint from any set $E^T_2$ of fact instances entailed by
$\langle R, W^{S,y}_{i-z} \rangle$, where ${z \geq x}$.
% This is because the timestamps of the instances ${E^T1}$ are a subset of those of
% the instances $W^{S,x}_i$, the timestamps of the instances ${E^T2}$ are a subset
% of those of the instances $W^{S,y}_{i-z}$, fact instances are distinct by timestamp and
% ${\{ t : f | t \in W^{S,x}_i\} \cap \{ t : f | t \in W^{S,y}_{i-z}\} = \varnothing}$.
\[ \{ e : \langle R , W^{S,x}_i \rangle \vdash e \} \cap \{ e : \langle R , W^{S,y}_{i-z} \rangle \vdash e \} = \varnothing \]
\end{axiom}

@@

A Continuous Datalog program (henceforth abbreviated C-Datalog)
$\Pi^C = \langle R, F^0, S, r \rangle$ is a tuple which comprises a
finite set of rules $R$, a finite set of static fact instances $F^0$
all of which have the instant $0$, a stream $S$ and a range ${r \in
\mathbb{N}}$ for the window over $S$.

@@

We consider the entailment of the C-Datalog program $\Pi^C$ not just
as a set of fact instances (as for TA-Datalog), but in three parts: a
set of fact instances $E^0$ entailed by the fact instances in $F^0$, a
stream $P$ of fact instances entailed by the fact instances in $F^0$
and S, and a stream $N$ of fact instance negations. $P$ and $N$ can be
considered as the continuous reasoning equivalents of the streams
produced by the \emph{IStream} and \emph{DStream} operators of the
\emph{CQL} continuous query language \citep{CQL}, respectively.

whose contents
are defined in 
Axioms~\ref{axiom:continuous datalog: positive window increment}, 
\ref{axiom:continuous datalog: negative window increment} and
\ref{axiom:continuous datalog: entailment precedes negation}. 
These specify that the streams 
\[ \Pi^C = \langle R, F, S, r \rangle \vDash \langle E, P, N \rangle \]
The truth value of a $\Pi^C$ is instantaneous, with the the truth
function of Continuous Datalog $T^{CD}$ taking both a program ${\Pi^C}$
and an instant ${i \in \mathbb{N}}$ as arguments.
\end{definition}

\begin{definition}[State of a $\Pi^C$]\label{def:continuous datalog: CDPt}
The state of a $\Pi^C$ at a given instant $i$ is the time-annotated
Datalog program ${\Pi^C_i = \langle R, F^0 \cup W^{S,r}_{i} \rangle}$,
where ${F^0 = \{ f | 0 : f \in F \}}$ is the set of persistent fact instances produced by
annotating each fact in $f$ in $F$ as becoming true at the first recognised instant.

>>>>

Continuous Datalog (henceforth C-Datalog) is an extension of positive
Datalog with stream-processing semantics, with a view to accommodating
the aims of the latter within the well-defined model-theoretic
semantics of the former, which are shared with RIF-Core
\cite{w3crifbld} and therefore support reasoning at the expressivity
of OWL 2 RL. As part of its application to Semantic Web reasoning, we
assume that the atomic formulae of the Datalog programs described
below correspond to RDF triples: $\langle subj, pred, obj\rangle$.

C-Datalog differs from conventional Datalog through the addition of
sources of continuously generated or reported facts (i.e streams), and
the application of sliding window semantics over those sources,
thereby providing for the interpretation of Continuous Datalog
programs as sequences of instantaneous states that are each valid
positive Datalog programs.  As such, these states each inherit the
semantics of positive Datalog.  Furthermore, it involves the
continuous derivation of entailments from this sequence of
instantaneous states as a set of streams of facts; these streams share
a data model with the streams of continuous queries, as defined by the
stream processing community, thereby allowing the integration of
Continuous Datalog programs into stream processing pipelines.

@@

\begin{definition}[Streams]
\labeldef{continuous datalog: stream}

A {\em stream} is an unbounded sequence of fact instances ordered by
their instants; the instants are non-decreasing in the stream, such
that for any two fact instances $\langle f_i, t_i \rangle$ and
$\langle f_j, t_j\rangle$ where the former appears on the stream
  before the latter, $t_i \leqslant t_j$.
\end{definition}

@@

From this description of instance entailment,
Axiom~\ref{axiom:continuous datalog: entailment disjointness} may be
derived:

\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
  If two sliding windows over the same stream $S$ cover non-overlapping time periods,
then a set $E^T_1$ of fact instances entailed by ${\langle R, W^{S,x}_i \rangle}$
is necessarily disjoint from any set $E^T_2$ of fact instances entailed by
$\langle R, W^{S,y}_{i-z} \rangle$, where ${z \geq x}$.
% This is because the timestamps of the instances ${E^T1}$ are a subset of those of
% the instances $W^{S,x}_i$, the timestamps of the instances ${E^T2}$ are a subset
% of those of the instances $W^{S,y}_{i-z}$, fact instances are distinct by timestamp and
% ${\{ t : f | t \in W^{S,x}_i\} \cap \{ t : f | t \in W^{S,y}_{i-z}\} = \varnothing}$.
\[
\{ e : \langle R , W^{S,x}_i \rangle \vdash e \} \cap \{ e : \langle R , W^{S,y}_{i-z} \rangle \vdash e \} = \varnothing
\]
\end{axiom}

@@

A Continuous Datalog program (henceforth abbreviated C-Datalog)
$\Pi^C = \langle R, F^0, S, r \rangle$ is a tuple which comprises a
finite set of rules $R$, a finite set of static fact instances $F^0$
(all of which have the instant $0$), a stream $S$ and a range
${r \in \mathbb{N}}$ for a window over $S$.

@@

We consider the entailment of the program $\Pi^C$ not just
as a set of fact instances (as for TA-Datalog), but in three parts: a
set of fact instances $E^0$ entailed by the fact instances in $F^0$, a
stream $P$ of fact instances entailed by the fact instances in $F^0$
and S, and a stream $N$ of fact instance negations: $\langle R, F^0,
S, r \rangle \models \langle E^0, P, N \rangle$. $P$ and $N$ can be
considered to be the continuous reasoning equivalents of the streams
produced by the \emph{IStream} and \emph{DStream} operators
respectively of the \emph{CQL} continuous query language \citep{CQL},
which create streams based on the additions to and deletions from a
relation.

In order to

\begin{definition}[Instanteous State of a C-Datalog program]
\label{def:continuous datalog: CDPt}

Given a C-Datalog program $\Pi^C$, we can

The truth value of a $\Pi^C$ is instantaneous, with the the truth
function of Continuous Datalog $T^{CD}$ taking both a program ${\Pi^C}$
and an instant ${i \in \mathbb{N}}$ as arguments.

The state of a $\Pi^C$ at a given instant $i$ is the time-annotated
Datalog program ${\Pi^C_i = \langle R, F^0 \cup W^{S,r}_{i} \rangle}$,
where ${F^0 = \{ f | 0 : f \in F \}}$ is the set of persistent fact instances produced by
annotating each fact in $f$ in $F$ as becoming true at the first recognised instant.
